A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-09-18 manuscript/paper.Rmd
update paper

\* Equal contribution  
¹ Data mining and Modelling for Biomedicine, VIB Center for Inflammation Research, Ghent, Belgium.  
² Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium.  
³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
⁴ Centre International de Recherche en Infectiologie, Inserm, U1111, Université Claude Bernard Lyon 1, CNRS, UMR5308, École Normale Supérieure de Lyon, Univ Lyon, F-69007, Lyon, France

```{r}
n_tools <- tools %>% filter(type == "algorithm") %>% nrow()
n_methods_evaluated <- nrow(filter(methods, evaluated, type %in% c("algorithm", "algorithm_test")))
n_control_methods_evaluated <- nrow(filter(methods, evaluated, type %in% c("control", "control_test")))
n_datasets_real <- 10
n_datasets_synthetic <- 20
```
# Abstract

@@

<!-- Evaluation overview -->
In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`), which go from the very basic topologies (linear, cyclical and bifurcating) to the most complex (cyclic graphs and disconnected graphs). →We defined several possible metrics to compare a prediction to the reference trajectory, and based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose a set of 4 metrics each assessing a different aspect of a trajectory: the cell ordering (`r label_metric("correlation")`), the topology (`r label_metric("him")`), the quality of differentially expressed features  (`r label_metric("featureimp_cor")`), and the quality of clustering of the cells into branches  (`r label_metric("F1_branches")`) (`r ref("fig", "ti_evaluation_overview", "d")`).←The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`). →For synthetic datasets we used several data simulators, including our own which simulates gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulates single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.←

@@

## Method stability and scalability

## Method quality control

@@

# Online Methods
## Data and code availability
→ 
The real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533)) [@cannoodt_goldstandardsinglecell_].

@@

<!-- Inclusion criterion -->
Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`. This resulted in the inclusion of  `r n_methods_evaluated` methods in the evaluation.

<!-- Components -->
The selected methods were dissected into a set of algorithmic components. Each component can have a significant impact on the performance, scalability, and output. Across all methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016]. Components are frequently shared between different algorithms. For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated.
### Method input
As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method.

<!-- Prior information -->
A large portion of the methods requires some form of prior information (e.g. a start cell) in order to be executable. Other methods allow to exploit certain prior information, though this is not required. Prior information can be supplied as a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states. Providing prior information to a TI method can be both a blessing and a curse. In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives. On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge. Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, restricting the applicability of such TI methods to well studied systems.

@@

* **Start cells** The identity of one or more start cells. For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) to each milestone with only outgoing edges. For ties, one random cell was chosen. For cyclic datasets, a random cell was chosen.
* **End cells** The identity of one or more end cells. Similar as the start cells, but now for every state with only ingoing edges.
* **# end states** Number of terminal states. Number of milestones with only ingoing edges.
* **Grouping** For each cell a label to which state/cluster/branch it belongs. For real data, the states from the gold/silver standard. For synthetic data, each milestone was seen as one group, and cells were assigned to their closest milestone.
* **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
* **Time course** For each cell a time point from which it was sampled. If available, directly extracted from the gold standard. For synthetic data: four timepoints were chosen, at which the cells were “sampled” to provide a time course information reflecting the one provided in real experiments.
### Common trajectory model
Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlap. We therefore created wrappers around each method (available at `r print_url("https://www.github.com/dynverse/dynmethods")`) and postprocessed its output into a common probabilistic trajectory model (`r ref("sfig", "overview_wrapping", "a")`). This model consists of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them. (ii) The milestone percentages contain, for each cell, its position between milestones, and sums for each cell to one. (iii) The trajectory model also allows for regions of delayed commitment, where cells are allowed to be positioned between three or more connected milestones. Regions must be explicitly defined in the trajectory model. Per region, one milestone must be directly connected to all other milestones in the network.

@@

```{r}
methods_evaluated <- read_rds(result_file("methods_evaluated.rds", "03-methods"))
# create functions to list the methods within a particular conversion category
conversion <- function(conversion_oi) {
  methods_evaluated %>% 
    filter(type %in% c("algorithm", "algorithm_test")) %>%
    filter(map_lgl(output, ~conversion_oi %in% .$outputs)) %>% 
    pull(name) %>% 
    sort() %>%
    label_vector()
}
```

* **Type 1, direct:** `r conversion("trajectory")`. The wrapped method directly returned a network of milestones, the regions of delayed commitment, and for each cell is given to what extent it belongs to a milestone. This usually indicates that additional transformations were required for the method not covered by any of the following output formats.
* **Type 2, linear pseudotime:** `r conversion("linear_trajectory")`. The method returned a pseudotime, which is translated into a linear trajectory where the milestone network contains two milestones and cells are positioned between these two milestones.
* **Type 3, cyclical pseudotime:** `r conversion("cyclic_trajectory")`. The method returned a pseudotime, which is translated into a cyclical trajectory where the milestone network contains three milestones and cells are positioned between these three milestones. →These milestones were positioned at pseudotime 0, ⅓ and ⅔ .←
* **Type 4, end state probability:** `r conversion("end_state_probabilities")`. The method returned a pseudotime and for each cell and end state a probability for how likely a cell will end up in a certain end state. →This was translated into a star-shaped milestone network, with one starting milestone ($M_0$) and several outer milestones ($M_i$) with regions of delayed commitment between all milestones. The milestone percentage of a cell to one of the outer milestones is equal to $\textrm{pseudotime} \times \textrm{Pr}_{M_i}$ , the milestone percentage to the starting milestone is equal to $1-\textrm{pseudotime}$ .←
* **Type 5, cluster assignment:** `r conversion("cluster_graph")`. The method returned a milestone network, and an assignment of each cell to a specific milestone. Cells were positioned onto the milestones they are assigned to.
* **Type 6, project onto nearest branch:** `r conversion("dimred_projection")`. The method returned a milestone network, and a dimensionality reduction of the cells and milestones. The cells were projected onto the closest nearest segment, thus determining the cells position along the milestone network. →If a method also returned a cluster assignment (type 5), we limited the projection of each cell to the closest edge connecting to the milestone of a cell. For these methods, we usually wrote two wrappers, one which included the projection, and one without.←
* →**Type 7, branch trajectory:** `r conversion("branch_trajectory")`. The method returned an assignment of cells to a particular branch, a pseudotime for each cell within the branch, and a network connecting these different branch. We converted this branch network to a milestone network by calculating the line graph of the network. The cells were then placed between each two milestones using the provided pseudotime. ←
* **Type 8, cell graph:** `r conversion("cell_graph")`. The method returned a network of cells, 
and which cell-cell transitions are part of the 'backbone' structure. Backbone cells with degree $\neq$ 2 were regarded as milestones, and all other cells were placed on transitions between the milestones. →If methods did not return a distance between pairs of cells, the cells were uniformly positioned between the two milestones. Otherwise, we first calculated the distance between two milestones as the sum of the distances between the cells, and then divided the distance of each pair of cells with the total distance to get the milestone percentages.←

@@

* **DPT:** We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to a tip point. This was then processed as described above.
* **Sincell:** To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.  This was then processed as described above.
* **SLICE:** As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. This local pseudotime was then processed as above.
* **MFA:** We used the branch assignment as state probabilities, which together with the global pseudotime were processed as described above.

The dynmethods github repository contains one vignette for each of the methods, discussing how to use the package to apply a method to a dataset, the specific implementation of the wrapper, and how the quality control scores of each TI method were derived.
→
### Control methods
We added several control positive and negative control methods: **random**, which generates a random network with 50 milestones according to a Barabási–Albert model, **identity**, which returns the gold standard, and **shuffle**, which will place the cells randomly on the gold standard milestone network.
### Off-the-shelf methods
For baseline performance, we added several "off-the-shelf" TI methods which can be run using just a couple of commands in R.←

* **Component 1:** This method returns the first component of a PCA dimensionality reduction as a linear trajectory. Although conceptually very simple, it has been used in certain studies [@kouno_temporaldynamicstranscriptional_2013; @zeng_pseudotemporalorderingsingle_2017].
* **Angle:** Similar to the previous method, this method computes the angle with respect to the origin in a two-dimensional PCA plot, and returns a cyclical trajectory.
→
* **Periodic PrinCurve:** Periodic principal curves, using the R princurve package
* **MST:** Dimensionality reduction with PCA, followed by clustering using the R mclust package, after which the clusters are connected using a minimal spanning tree. The cells are then projected onto the edges of the MST.←
## Real datasets
We gathered `r sum(list_datasets()$source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`). The scripts to download and process these datasets are available on our repository ([https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets](https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets)). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalised and filtered using a common pipeline, discussed later.

@@

The workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment. At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable. For every synthetic dataset, we used a random real dataset as a reference dataset (from those described earlier), making sure the number of variable genes and cells were similar.

#### Network generation

@@

### dyntoy

@@

### Splatter
## Dataset filtering and normalisation
We used a standard single-cell RNA-seq preprocessing pipeline which applies parts of the scran and scater Bioconductor packages [@lun_stepbystepworkflowlowlevel_2016]. The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed using median absolute deviations (set to 3). We required that a gene was expressed in at least 5% of the cells, and that it should have an average expression higher than 0.05. Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5% and a biological component higher than 0.5. As a final filter, we removed both zero genes and cells until convergence.
## Evaluation metrics
The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011]. Furthermore, We defined three metrics for comparing the similarity of predicted trajectories to a gold standard (`r ref("fig", "ti_evaluation_overview", "c")`). Each metric assesses the performance of a different aspect of the trajectory (`r ref("sfig", "score_aspects")`); (i) the correlation metric measures the similarity in pairwise cell-cell distances; (ii) the RF MSE metric assesses whether cell neighbourhoods are similar in both trajectories; and (iii) the edge flip scores assesses similarity in trajectory topologies.

@@

>>>>

\* Equal contribution  
¹ Data mining and Modelling for Biomedicine, VIB Center for Inflammation Research, Ghent, Belgium.  
² Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium.  
³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
⁴ Centre International de Recherche en Infectiologie, Inserm, U1111, Université Claude Bernard Lyon 1, CNRS, UMR5308, École Normale Supérieure de Lyon, Univ Lyon, F-69007, Lyon, France
```{r}
n_tools <- tools %>% filter(type == "algorithm") %>% nrow()
n_methods_evaluated <- nrow(filter(methods, evaluated, type %in% c("algorithm", "algorithm_test")))
n_control_methods_evaluated <- nrow(filter(methods, evaluated, type %in% c("control", "control_test")))
n_datasets_real <- 10
n_datasets_synthetic <- 20
```
# Abstract

@@

<!-- Evaluation overview -->
In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`), which go from the very basic topologies (linear, cyclical and bifurcating) to the most complex (cyclic graphs and disconnected graphs). →We defined several possible metrics to compare a prediction to the reference trajectory (`r ref("snote", "metrics")`). Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose 4 metrics each assessing a different aspect of a trajectory: the cell ordering (`r label_metric("correlation")`), the topology (`r label_metric("him")`), the quality of differentially expressed features  (`r label_metric("featureimp_cor")`), and the quality of clustering of the cells into branches  (`r label_metric("F1_branches")`) (`r ref("fig", "ti_evaluation_overview", "d")`).←The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`). →For synthetic datasets we used several data simulators, including our own which simulates gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011]. We paired each simulated datasets with a real dataset to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties.←

@@

## Method stability and scalability
## Method quality control

@@

________________
# Online Methods
## Data and code availability
→ 
The real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533)) [@cannoodt_goldstandardsinglecell_].

@@

<!-- Inclusion criterion -->
Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`. The discussions on individual methods can be found at `r print_url("https://github.com/dynverse/dynmethods/issues?q=label:unwrappable")` This resulted in the inclusion of  `r n_methods_evaluated` methods in the evaluation.

<!-- Components -->
The selected methods were dissected into a set of algorithmic components. Each component can have a significant impact on the performance, scalability, and output. Across all methods, these components can be broadly grouped into two stages; (i) conversion to a simplified representation using dimensionality reduction, clustering or graph building and (ii) ordering the cells along the simplified representation [@cannoodt_computational_2016]. Components are frequently shared between different algorithms. For example, minimal spanning trees (MST), used in the first single-cell RNA-seq trajectory inference methods [@trapnell_dynamicsregulatorscell_2014], is shared by almost half of the methods we evaluated.

→
### Method wrappers
Each method was wrapped within a docker and singularity container (available at `r print_url("https://github.com/dynverse/dynverse/tree/master/methods")`). These containers are automatically built on both Singularity Hub (`r print_url("https://singularity-hub.org/")`) and Docker Hub (`r print_url("https://hub.docker.com/u/dynverse")`).
For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors. This script will read in the input data (as described below), run the method, and output the files required to construct a trajectory (as described below). To wrap each method, we based ourselves on example scripts or tutorials which are mentioned in the comments of each wrapper script. We also created a script to generate an example dataset, which is used to automatically test every method using travis continuous integration (`r print_url("https://travis-ci.org/dynverse")`).

We used the github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata, and the quality control. These discussion can be viewed on github (`r print_url('https://github.com/dynverse/dynmethods/issues?q=label:"method+discussion"')`). About ⅓ of authors responded, and we improved the wrappers based on their feedback. 
←
### Method input
As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method.
<!-- Prior information -->
A large portion of the methods requires some form of prior information (e.g. a start cell) in order to be executable. Other methods optionally allow to exploit certain prior information, though this is not required. Prior information can be supplied as a starting cell from which the trajectory will originate, a set of important marker genes, or even a grouping of cells into cell states. Providing prior information to a TI method can be both a blessing and a curse. In one way, prior information can help the method to find the correct trajectory among many, equally likely, alternatives. On the other hand, incorrect or noisy prior information can bias the trajectory towards current knowledge. Moreover, prior information is not always easily available, and its subjectivity can therefore lead to multiple equally plausible solutions, restricting the applicability of such TI methods to well studied systems.

@@

* **Start cells** The identity of one or more start cells. For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) to each milestone with only outgoing edges. For ties, one random cell was chosen. For cyclic datasets, a random cell was chosen.
* **End cells** The identity of one or more end cells. Similar as the start cells, but now for every state with only ingoing edges.
* **# end states** Number of terminal states. Number of milestones with only ingoing edges.
* **Grouping** For each cell a label to which state/cluster/branch it belongs. For real data, the states from the gold/silver standard. For synthetic data, each milestone was seen as one group, and cells were assigned to their closest milestone.
* **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
→
* **Discrete time course** For each cell a time point from which it was sampled. If available, directly extracted from the gold standard, otherwise the geodesic distance from the root milestone was used. For synthetic data, four discrete timepoints were chosen, at which the cells were “sampled” to provide a time course information reflecting the one provided in real experiments.
* **Continuous time course** For each cell a time point from which it was sampled. For real data this was equal to the discrete time course, for synthetic data we used the internal simulation time of each simulator.
←
### Common trajectory model
Due to the absence of a common format for trajectory models, most methods return a unique set of output formats with few overlaps. →We therefore post-processed the output of each method into a common probabilistic trajectory model← (`r ref("sfig", "overview_wrapping", "a")`). This model consisted of three parts: (i) The milestone network represents the overall network topology, and contains edges between different milestones and the length of the edge between them. (ii) The milestone percentages contain, for each cell, its position between milestones, and sums for each cell to one. (iii) The trajectory model also allows for regions of delayed commitment, where cells are allowed to be positioned between three or more connected milestones. Regions must be explicitly defined in the trajectory model. Per region, one milestone must be directly connected to all other milestones →of the region.←

@@

```{r}
methods_evaluated <- read_rds(result_file("methods_evaluated.rds", "03-methods"))
special_methods <- c("mfa", "dpt", "slice", "sincell", "calista", "urd")
# create functions to list the methods within a particular conversion category
conversion <- function(conversion_oi) {
  methods_evaluated %>% 
    filter(type %in% c("algorithm", "algorithm_test")) %>%
    filter(map_lgl(output, ~conversion_oi %in% .$outputs)) %>% 
    mutate(name = paste0(name, ifelse(id %in% special_methods, "\\*", ""))) %>% 
    pull(name) %>% 
    sort() %>%
    label_vector()
}
```

* **Type 1, direct:** `r conversion("trajectory")`. The wrapped method directly returned a network of milestones, the regions of delayed commitment, and for each cell is given to what extent it belongs to a milestone. →In some cases, this← indicates that additional transformations were required for the method not covered by any of the following output formats.
* **Type 2, linear pseudotime:** `r conversion("linear_trajectory")`. The method returned a pseudotime, which is translated into a linear trajectory where the milestone network contains two milestones and cells are positioned between these two milestones.
* **Type 3, cyclical pseudotime:** `r conversion("cyclic_trajectory")`. The method returned a pseudotime, which is translated into a cyclical trajectory where the milestone network contains three milestones and cells are positioned between these three milestones. →These milestones were positioned at pseudotime 0, ⅓ and ⅔ .←
* **Type 4, end state probability:** `r conversion("end_state_probabilities")`. The method returned a pseudotime and for each cell and end state a probability for how likely a cell will end up in a certain end state. →This was translated into a star-shaped milestone network, with one starting milestone ($M_0$) and several outer milestones ($M_i$) with regions of delayed commitment between all milestones. The milestone percentage of a cell to one of the outer milestones was equal to $\textrm{pseudotime} \times \textrm{Pr}_{M_i}$ , the milestone percentage to the starting milestone was equal to $1-\textrm{pseudotime}$ .←
* **Type 5, cluster assignment:** `r conversion("cluster_graph")`. The method returned a milestone network, and an assignment of each cell to a specific milestone. Cells were positioned onto the milestones they are assigned to.
* **Type 6, project onto nearest branch:** `r conversion("dimred_projection")`. The method returned a milestone network, and a dimensionality reduction of the cells and milestones. The cells were projected onto the closest nearest segment, thus determining the cells position along the milestone network. →If a method also returned a cluster assignment (type 5), we limited the projection of each cell to the closest edge connecting to the milestone of a cell. For these methods, we usually wrote two wrappers, one which included the projection, and one without.←
* →**Type 7, branch trajectory[a]:** `r conversion("branch_trajectory")`. The method returned an assignment of cells to a particular branch, a pseudotime for each cell within the branch, and a network connecting these different branch. We converted this branch network to a milestone network by calculating the line graph of the network. The cells were then placed between each two milestones using the provided pseudotime. ←
* **Type 8, cell graph:** `r conversion("cell_graph")`. The method returned a network of cells, 
and which cell-cell transitions are part of the 'backbone' structure. Backbone cells with degree $\neq$ 2 were regarded as milestones, and all other cells were placed on transitions between the milestones. →If methods did not return a distance between pairs of cells, the cells were uniformly positioned between the two milestones. Otherwise, we first calculated the distance between two milestones as the sum of the distances between the cells, and then divided the distance of each pair of cells with the total distance to get the milestone percentages.←

@@

* **DPT:** We projected the cells onto the cluster network, consisting of a central milestone (this cluster contains the cells which were assigned to the "unknown" branch) and three terminal milestones, each corresponding to a tip point. →This was then processed as a type 1, direct,  method.←
* **Sincell:** To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, default at 25%.  →This was then processed as a type 8, cell graph, method.←
* **SLICE:** As discussed in the vignette of SLICE, we ran principal curves one by one for every edge detected by SLICE. →This was then processed as a type 1, direct, method.←
* **mfa:** We used the branch assignment as state probabilities, which together with the global pseudotime were processed →as a type 4, end state probabilities, method.←
→
* **CALISTA:** We assigned the cells to the branch at which the sum of the cluster probabilities of two connected milestones was the highest. The cluster probabilities of the two selected milestones were then used as milestone percentages. This was then processed as a type 1, direct, method.
* **URD:** We extracted the pseudotime of a cell within each branch using the y positions in the tree layout. This was then further processed as a type 1, direct, method.

More information on how each method was wrapped can be found within the comments of each wrapper script.
←

→
### Control methods
We added several control positive and negative control methods: **random**, which generates a random network with 50 milestones according to a Barabási–Albert model, **identity**, which returns the gold standard, and **shuffle**, which will place the cells randomly on the gold standard milestone network.
### Off-the-shelf methods
For baseline performance, we added several "off-the-shelf" TI methods which can be run using just a couple of commands in R.←

* **Component 1:** This method returns the first component of a PCA dimensionality reduction as a linear trajectory. Although conceptually very simple, it has been used in certain studies [@kouno_temporaldynamicstranscriptional_2013; @zeng_pseudotemporalorderingsingle_2017].
* **Angle:** Similar to the previous method, this method computes the angle with respect to the origin in a two-dimensional PCA, and uses this angle as pseudotime for a cyclical trajectory.
→
* **Periodic PrinCurve:** Periodic principal curves, using the R princurve package
* **MST:** Dimensionality reduction with PCA, followed by clustering using the R mclust package, after which the clusters are connected using a minimal spanning tree. The cells are then projected onto the edges of the MST.
←

## Real datasets
We gathered `r sum(list_datasets()$source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`). The scripts to download and process these datasets are available on our repository ([https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets](https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets)). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalised and filtered using a common pipeline, discussed later.

@@

The workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment. At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable. For every synthetic dataset, we used a random real dataset as a reference dataset (from those described earlier), making sure the number of variable genes and cells were similar.
#### Network generation

@@

→
### dyntoy

@@

### Splatter
←
## Dataset filtering and normalisation
We used a standard single-cell RNA-seq preprocessing pipeline which applies parts of the scran and scater Bioconductor packages [@lun_stepbystepworkflowlowlevel_2016]. The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed using median absolute deviations (set to 3). We required that a gene was expressed in at least 5% of the cells, and that it should have an average expression higher than 0.05. Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5% and a biological component higher than 0.5. As a final filter, we removed both zero genes and cells until convergence.
## Evaluation metrics
The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011]. →Furthermore, a trajectory is a model with multiple layers of complexity, which calls for several metrics each assessing a different layer. We therefore defined several possible metrics for comparing trajectories, each investigating different layers. These are all discussed in `r ref("snote", "metrics")` along with examples and robustness analyses when appropriate.

Next, we created a set of rules to which we think a good trajectory metric should conform, and tested this empirically for each metric by comparing scores before and after perturbing a dataset (`r ref("snote", "metrics")`). Based on this analysis, we chose 4 metrics for ←

  We defined three metrics for comparing the similarity of predicted trajectories to a gold standard (`r ref("fig", "ti_evaluation_overview", "c")`). Each metric assesses the performance of a different aspect of the trajectory (`r ref("sfig", "score_aspects")`); (i) the correlation metric measures the similarity in pairwise cell-cell distances; (ii) the RF MSE metric assesses whether cell neighbourhoods are similar in both trajectories; and (iii) the edge flip scores assesses similarity in trajectory topologies.

@@

[a]Gotta update the figure to reflect this
